Density-dependent selection and the limits of relative fitness

### 2018-03-09 manuscript.tex
l

We now outline the properties of selection in the variable-density lottery, starting with selection on competitive ability $c$. Eq.~\eqref{eq:Nmaster} shows that $c$ is not a density-regulating trait: $c$ only affects the relative likelihood for each type to win a contested territory, not whether a territory is contested in the first place. Selection between types which only differ in $c$ occurs without changing density $N$. However, $c$-selection is density-dependent, with the strength of selection (measured by the difference between types in the per-capita growth rate $\Delta n_i/n_i$) peaking at an intermediate density (Fig.~\ref{fig:DDS_lottery}). This intermediate peak occurs because most territories are claimed without contest at low density, whereas few unoccupied territories are available to be contested at high density.

Turning to selection on the density-regulating traits $b$ and $d$, we write Eq.~\eqref{eq:masterequalc} in the form
\begin{equation}
\frac{\Delta n_i}{n_i} = \frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N)-d_i, \label{eq:bdensitydependence}
\end{equation}
where we have used that fact that $L=\overline{b}N/T$, and $\overline{b}$ is the population mean $b$. It is clear from Eq.~\eqref{eq:bdensitydependence} that selection on adult mortality $d$ is independent of density. On the other hand, $b$-selection is density-dependent. The advantage of having greater $b$ declines with density because fewer territories are available to to be claimed (Fig.~\ref{fig:DDS_lottery}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{DDS_lottery.pdf}
\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as difference in propagule success rate $\Delta n_j/m_j-\Delta n_i/n_i$ between an adaptive variant $j$ and a wildtype $i$ present at the same frequency. Here $b_i=1$, $d_i=0.5$ and $c_i=1$. For $b$-selection we set $b_j=b_i(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$. $d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as  available territories become scarce. The equilibrium density for the $i$ type is $\approx 0.4$.}
\end{figure}

@@

We now start evaluating the limits of Eq.~\eqref{eq:canonical} in the context of the variable-density lottery model, focusing in this section on the issue of whether density varies as a result of selection \citep{prout_1980}. We showed in the previous section that $c$-selection in the variable-density lottery does not affect density. 
However, this is not the case in most previous models of density-dependent selection, which have been strongly influenced by the notion of ``$K$-selection'' and the associated questions about whether selection in crowded populations will tend to increase population density \citep{anderson_1971,roughgarden_1979,lande_2009}. To clarify the connection between ``$K$-selection'' and the variable density lottery model, we thus revisit MacArthur's analysis of selection in crowded populations \citep{macarthur_1967}.

MacArthur considers a population with two types that have densities $n_1$ and $n_2$ subject to density-dependent growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
The environment is assumed to remain constant apart from the type densities. The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources. This defines nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

To formalize the relationship between nullclines, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$. These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines. Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; following widespread convention, we will refer to selection on these saturation densities as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a). The other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types. For instance, in the Lotka-Volterra competition model we have
\begin{align}
f_1(n_1,n_2) = r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1\nonumber\\
f_2(n_1,n_2) = r_2(1-\alpha_{22}n_1-\alpha_{21}n_2)n_2\label{eq:LV}
\end{align}
where $\alpha_{11}=1/K_{11}$ and $\alpha_{22}=1/K_{22}$ measure competitive effects within types, while $\alpha_{12}=1/K_{12}$ and $\alpha_{21}=1/K_{21}$ measure competitive effects between types (Fig.~\ref{fig:LVvslottery}a).

In summary, MacArthur's conclusion that  ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967} means that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}. This general idea is much broader than ``$K$-selection'' in the sense of selection for greater saturation density, and applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

@@

It is obvious in the Lotka-Volterra model that selection can favor a superior competitor in a crowded population even if its saturation density is lower than that of the other types present. Nevertheless, the Lotka-Volterra model still couples population density to selection \citep{smouse_1976}. Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$). Even though the initial and final densities of a sweep are the same, density is not constant over a sweep (constant density is possible but only occurs for a highly restricted subset of $r$ and $\alpha$ values; further details in Appendix C; also see \citealt{mallet_2012}). Intuitively, for one type to exclude another with the same saturation density, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing a dip in $N$ over the sweep.

>>>>

We now outline the basic properties of selection on $b$, $c$ and $d$ in the variable-density lottery.

Eq.~\eqref{eq:Nmaster} shows that $c$ is not a density-regulating trait: $c$ only affects the relative likelihood for each type to win a contested territory, not the number of territories that are contested in the first place. Thus, selection between types which only differ in $c$ occurs without causing $N$ to change. However, $c$-selection is density-dependent, with the strength of selection (measured by the difference between types in the per-capita growth rate $\Delta n_i/n_i$) peaking at an intermediate density (Fig.~\ref{fig:DDS_lottery}). This intermediate peak occurs because at low density most territories are claimed without contest, whereas at high density few unoccupied territories are available to be contested.

The birth and mortality rates $b$ and $d$ do regulate density. To see how selection on $b$ and $d$ depends on density, we write Eq.~\eqref{eq:masterequalc} in the form
\begin{equation}
\frac{\Delta n_i}{n_i} = \frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N)-d_i, \label{eq:bdensitydependence}
\end{equation}
where we have used that fact that $L=\overline{b}N/T$, and $\overline{b}$ is the population mean $b$. It is clear from Eq.~\eqref{eq:bdensitydependence} that selection on adult mortality $d$ is independent of density. On the other hand, $b$-selection is density-dependent; the advantage of having greater $b$ declines with density because fewer territories are available to to be claimed (Fig.~\ref{fig:DDS_lottery}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{DDS_lottery.pdf}
\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as the difference in propagule success rate $\Delta n_j/m_j-\Delta n_i/n_i$ between an adaptive variant $j$ and a wildtype $i$ present at the same frequency. Here $b_i=1$, $d_i=0.5$ and $c_i=1$. For $b$-selection we set $b_j=b_i(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$. $d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as  available territories become scarce. The equilibrium density for the $i$ type is $\approx 0.4$.}
\end{figure}

@@

Equipped with an understanding of selection in the variable-density lottery model, we now turn to the broader issue of how density responds to selection \citep{prout_1980}. Going back to some of the earliest work on the topic \citep{kostitzin_1939,christiansen_2004}, most models of selection in density-regulated populations have some form of ``$K$-selection'' in the sense that selection in crowded populations increases saturation density. This makes sense intuitively because births balance deaths at equilibrium, and adaptive mutants must overcome the existing balance to expand. Yet, as we saw in the previous section, $c$-selection has no effect on population density in the variable-density lottery model. To make sense of how this property of $c$-selection fits with more familar population growth models, we revisit MacArthur's analysis of $K$-selection \citep{macarthur_1967}.

MacArthur considered a population with two types that have densities $n_1$ and $n_2$ subject to density-dependent growth,
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
The environment is assumed to remain constant apart from changing type densities. The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because the resources required for growth are limited. This defines nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. Thus, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

MacArthur used ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(K_{21},0)=0$ and $f_2(0,K_{22})=0$. These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines. Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model (Fig.~\ref{fig:Ksel}a). The other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types. To be more concrete, in the Lotka-Volterra competition model we have
\begin{align}
f_1(n_1,n_2) = r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1\nonumber\\
f_2(n_1,n_2) = r_2(1-\alpha_{22}n_1-\alpha_{21}n_2)n_2\label{eq:LV}
\end{align}
where $\alpha_{11}=1/K_{11}$ and $\alpha_{22}=1/K_{22}$ measure competitive effects within types, while $\alpha_{12}=1/K_{12}$ and $\alpha_{21}=1/K_{21}$ measure competitive effects between types. Hence, ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967} in the sense that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}. This general idea is much broader than ``$K$-selection'' in the sense of selection for greater saturation density.

@@

The Lotka-Volterra competition model makes it is clear that selection in crowded populations need not favor ever greater saturation density \citep{gill_1974}. For instance, Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$).

Nevertheless, the Lotka-Volterra model still couples population density to selection \citep{smouse_1976}.  Even though the initial and final densities of a sweep are the same, density is not constant over a sweep (constant density is possible but only occurs for a highly restricted subset of $r$ and $\alpha$ values; further details in Appendix C; also see \citealt{mallet_2012}). Intuitively, for one type to exclude another with the same saturation density, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing a dip in $N$ over the sweep.

(Fig.~\ref{fig:LVvslottery}a)
